---
title: 'Day 1, Session 2: Data wrangling'
author: "Dusica Filipovic Durdevic"
date: "2026-09-14"
output:
  revealjs::revealjs_presentation:
    theme: simple
    highlight: pygments
    transition: none
    center: false
    css: custom.css
    slide_level: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Scripting preserves future efforts

- Saving the code

  - Saving time
  
  - Data reproducibility
  


## Loading the data 

- Opening data file from the working directory

```{r, warning = FALSE, message = FALSE}
read.table("Data_practice.txt",sep="\t", header=TRUE)
```

- Loading the data / opening data file from the target directory using path

```{r, warning = FALSE, message = FALSE}
read.table("c:/Eowyn/Dokumenti/Obuke/!MOJE_2026/Podaci_vezba.txt",sep="\t", header=TRUE)
```

- Working with large data files, we will assign them a name

```{r, warning = FALSE, message = FALSE}
df <- read.table("Data_practice.txt",sep="\t", header=TRUE)
```

```{r, warning = FALSE, message = FALSE}
df <- read.csv("Data_practice.csv")           # comma delimited (default)
```

```{r, warning = FALSE, message = FALSE}
df <- read.delim("Data_practice.txt")         # tab delimited (default)
```

```{r, warning = FALSE, message = FALSE}
# library(readxl)
# df <- read_excel("Data_practice.xlsx")
```

> Fun fact

> You will come up with a name for a data frame yourself, but a common practice is to name it df

> Caution

> Avoid reserved names (e.g. data)


## Data wrangling

## Data wrangling

- A set of procedures that we apply to a data frame before the analysis

  - Cleaning the data
    - Excluding participants, items
    - Filtering
    
  - Transforming variables
    - log-transforming, scaling
    
  - Reshaping the data frame
    - Wide-to-long format
    - Joining dataframes
  
  - Renaming 
    - Variables
    - Category levels
  
## Let's start

- Load the data

```{r, warning = FALSE, message = FALSE}
df <- read.table("Data_practice.txt",sep="\t", header=TRUE)
```

- Inspect the number of rows and columns of the data frame

```{r, warning = FALSE, message = FALSE}
dim(df)
```


## Cleaning the data

## Load and inspect the data

- Have a quick look at the data frame

```{r, warning = FALSE, message = FALSE}
head(df)
```

- Inspect the type of variables

```{r, warning = FALSE, message = FALSE}
str(df)
```

## Conduct conversion to appropriate data type

  - Base R

```{r, warning = FALSE, message = FALSE}
df$Participant = as.factor(df$Participant)
df$Manipulation = as.factor(df$Manipulation)
df$DominantHand = as.factor(df$DominantHand )
df$Recall = as.numeric(df$Recall)
df$IQ = as.numeric(df$IQ)
```

  - dplyr

```{r, warning = FALSE, message = FALSE}
library(dplyr)

df = df %>%
  mutate(
    Participant  = as.factor(Participant),
    Manipulation = as.factor(Manipulation),
    DominantHand = as.factor(DominantHand),
    Recall       = as.numeric(Recall),
    IQ           = as.numeric(IQ)
  )
```

- Inspect the type of variables again

```{r, warning = FALSE, message = FALSE}
str(df)
```

## Duplicated rows

- Base R

  - Check if there are duplicated rows
  
```{r, warning = FALSE, message = FALSE}
any(duplicated(df))
```

  - Count duplicated rows

```{r, warning = FALSE, message = FALSE}
sum(duplicated(df))
```

  - Show duplicated rows

```{r, warning = FALSE, message = FALSE}
df[duplicated(df), ]
```

  - Remove duplicated rows

```{r, warning = FALSE, message = FALSE}
df = df[!duplicated(df), ]  
```

- dplyr

```{r, warning = FALSE, message = FALSE}
df %>%
  filter(duplicated(.)) 
```

  - Remove duplicated rows

```{r, warning = FALSE, message = FALSE}
df %>%
  distinct() 
```

## Missing values

- Base R

  - Check if there are any missing values
  
```{r, warning = FALSE, message = FALSE}
# in total
sum(is.na(df))
# per column
colSums(is.na(df)) 
```

  - Remove rows with missing values
  
```{r, warning = FALSE, message = FALSE}
# remove rows with any missing value
df_clean = na.omit(df)
# remove rows with missing values in specific columns
df_clean = df[complete.cases(df[, c("IQ", "DominantHand")]), ]
```

  - Replace missing values with group mean
  
```{r, warning = FALSE, message = FALSE}
df$IQ[is.na(df$IQ)] = mean(df$IQ, na.rm = TRUE)
# na.rm = TRUE tells R to ignore missing values 
#              when calculating the mean
```

- dplyr, tidyr

  - Check if there are any missing values
  
```{r, warning = FALSE, message = FALSE}
df %>% 
  summarise(across(everything(), ~ sum(is.na(.))))
```

  - Remove rows with missing values
  
```{r, warning = FALSE, message = FALSE}
library(tidyr)

# remove rows with any missing value
df_clean = df %>% 
  drop_na() 
# remove rows with missing values in specific columns
df_clean = df %>% 
  drop_na(IQ, DominantHand)
```

  - Replace missing values with group mean
  
```{r, warning = FALSE, message = FALSE}
df = df %>%
  mutate(IQ = if_else(is.na(IQ), mean(IQ, na.rm = TRUE), IQ))
```

## Inspecting levels of categorical variables

- Base R

  - See the levels

```{r, warning = FALSE, message = FALSE}
levels(df$Manipulation)
```

  - Frequency table (count per level)

```{r, warning = FALSE, message = FALSE}
table(df$Manipulation)
```

  - Number of levels

```{r, warning = FALSE, message = FALSE}
nlevels(df$Manipulation)
```

  - If a variable is not a factor

```{r, warning = FALSE, message = FALSE}
unique(df$Manipulation)
```

  - Crosstabulaton (contingency tables)

```{r, warning = FALSE, message = FALSE}
table(df$Manipulation, df$DominantHand)
```

```{r, warning = FALSE, message = FALSE}
xtabs(~Manipulation + DominantHand, data= df)
```

> Fun fact

> ~ reads as "depending of" or "as a function of"


- dplyr

  - See the levels

```{r, warning = FALSE, message = FALSE}
df %>% 
  distinct(Manipulation)
```

  - Frequency table (count per level)

```{r, warning = FALSE, message = FALSE}
df %>% 
  count(Manipulation)
```

  - Number of levels

```{r, warning = FALSE, message = FALSE}
n_distinct(df$Manipulation)
```

  - Crosstabulaton (contingency tables)

```{r, warning = FALSE, message = FALSE}
df %>% 
  count(Manipulation, DominantHand)
```

  - Percentages
  
```{r, warning = FALSE, message = FALSE}
df %>%
  count(Manipulation) %>%
  mutate(pct = n / sum(n) * 100)
```

## Recoding variables

- Base R

  - Creating new variables

```{r, warning = FALSE, message = FALSE}
df$PercRecall = 100 * df$Recall / max(df$Recall)
df$PercRecall
```

  - Rounding values

```{r, warning = FALSE, message = FALSE}
df$PercRecall = round(df$PercRecall,2)
df$PercRecall
```

  - Combining levels

```{r, warning = FALSE, message = FALSE}
df$ManipHand = as.factor(
                    paste(df$Manipulation, df$DominantHand, sep = "_"))
levels(df$ManipHand)
```

  - Creating new categorical variable based on median split

```{r, warning = FALSE, message = FALSE}
df$IQ_cat_med = as.factor(
                    ifelse(df$IQ > median(df$IQ, na.rm = TRUE), 
                                   "highIQ", "lowIQ"))
levels(df$IQ_cat_med)
```

  - Creating new categorical variable based on quantile split

```{r, warning = FALSE, message = FALSE}
df$IQ_cat_qu = as.factor(
                      cut(df$IQ, 
                      breaks = quantile(df$IQ, probs = seq(0, 1, 0.25), 
                                        na.rm = TRUE),
                      labels = c("Q1", "Q2", "Q3", "Q4"),
                      include.lowest = TRUE)) # technical feature,
                                  # otherwise they will become NA
levels(df$IQ_cat_qu)
table(df$IQ_cat_qu)
```

  - Creating new categorical variable by recoding IQ_cat_qu levels to average (Q2, Q3) and extreme (Q1, Q4)

```{r, warning = FALSE, message = FALSE}
df$IQ_extreme = as.factor(
          ifelse(df$IQ_cat_qu %in% c("Q2", "Q3"), "average", "extreme"))
levels(df$IQ_extreme)
table(df$IQ_extreme)
```

- dplyr

  - Creating new variables

```{r, warning = FALSE, message = FALSE}
df = df %>%
  mutate(PercRecall = 100 * Recall / max(Recall, na.rm = TRUE))
df$PercRecall
```

  - Rounding values

```{r, warning = FALSE, message = FALSE}
df = df %>%
  mutate(PercRecall = round(PercRecall, 2))
df$PercRecall
```

  - Combining levels

```{r, warning = FALSE, message = FALSE}
df = df %>%
  mutate(ManipHand = as.factor(
                        paste(Manipulation, DominantHand, sep = "_")))
levels(df$ManipHand)
```

  - Creating new categorical variable based on median split

```{r, warning = FALSE, message = FALSE}
df = df %>%
  mutate(IQ_cat_med = as.factor(
                        if_else(IQ > median(IQ, na.rm = TRUE), 
                                "highIQ", "lowIQ")))
levels(df$IQ_cat_med)
```

  - Creating new categorical variable based on quantile split

```{r, warning = FALSE, message = FALSE}
df = df %>%
  mutate(IQ_cat_qu = as.factor(cut(IQ,
                               breaks = quantile(IQ, probs = seq(0, 1, 0.25), 
                                                            na.rm = TRUE),
                               labels = c("Q1", "Q2", "Q3", "Q4"),
                                                   include.lowest = TRUE)))
levels(df$IQ_cat_qu)
table(df$IQ_cat_qu)
```

  - Creating new categorical variable by recoding IQ_cat_qu levels to average (Q2, Q3) and extreme (Q1, Q4)

```{r, warning = FALSE, message = FALSE}
df = df %>%
  mutate(IQ_extreme = as.factor(if_else(IQ_cat_qu %in% c("Q2", "Q3"), 
                                              "average", "extreme")))
levels(df$IQ_extreme)
table(df$IQ_extreme)
```




## Other fun stuff

- Sorting the data

```{r, warning = FALSE, message = FALSE}
df[order(df$Recall),] 
```

## Calculating the sum of the column values

```{r, warning = FALSE, message = FALSE}
sum(df$Recall)
```

## Calculating the mean of the column values

```{r, warning = FALSE, message = FALSE}
mean(df$Recall)
```

## Calculating standard deviation of the column values

```{r, warning = FALSE, message = FALSE}
sd(df$Recall)
```
 
## Calculating the mean separately for factor levels

```{r, warning = FALSE, message = FALSE}
mean(df[df$Manipulation=="control_group",]$Recall) 
```

```{r, warning = FALSE, message = FALSE}
mean(df[df$Manipulation=="experimental_group",]$Recall) 
```

## Calculating the mean separately for multiple levels at once

```{r, warning = FALSE, message = FALSE}
tapply(df$Recall, df$Manipulation, mean) 
```

```{r, warning = FALSE, message = FALSE}
with(df, tapply(Recall, Manipulation, mean)) 
```

## Calculating means by crossing factor levels

```{r, warning = FALSE, message = FALSE}
with(df, tapply(Recall, list(Manipulation, DominantHand), mean))
```

## Descriptive statistics at once

```{r, warning = FALSE, message = FALSE}
summary(df)
```

```{r, warning = FALSE, message = FALSE}
psych::describe(df)
```

> Caution

> psych::describe() calculates descriptives for all variables, even categorical (using internal numerical values)

  - Make sure to select only numerical variables when using psych::describe()
  
```{r, warning = FALSE, message = FALSE}
df %>% select(where(is.numeric)) %>% psych::describe()
df %>% select(where(is.factor)) %>% summary()
```

## Going wild and running a t-test while we're at it :)

```{r, warning = FALSE, message = FALSE}
t.test(Recall ~ Manipulation, data=df)
```

## Correlation

```{r, warning = FALSE, message = FALSE}
cor.test(df$IQ, df$Recall)
```


## LET'S WRANGLE REAL DATA

- Today

  - Using Base R

- Tommorow

  - Using dplyr
    - Session 1: step by step
    - Session 2: pipeline

## Load a small (but real) dataset

```{r, warning = FALSE, message = FALSE}
df1=read.table("Dusica.nouns.2004.txt",sep="\t", header = TRUE)
```

- Inspect the number of rows and columns of the data frame

```{r, warning = FALSE, message = FALSE}
dim(df1)
```

## Cleaning the data

- Have a quick look at the data frame

```{r, warning = FALSE, message = FALSE}
head(df1)
```

- Have a quick look at the column names

```{r, warning = FALSE, message = FALSE}
colnames(df1)
```

- Inspect the type of variables

```{r, warning = FALSE, message = FALSE}
str(df1)
```

## Conduct conversion to appropriate data type

  - Base R

```{r, warning = FALSE, message = FALSE}
df1$Participant = as.factor(df1$Participant)
df1$Lexicality = as.factor(df1$Lexicality)
df1$PolyLevelCategory  = as.factor(df1$PolyLevelCategory )
df1$Word = as.factor(df1$Word)
df1$FreCategory = as.factor(df1$FreCategory)
df1$PartSpeedCategory = as.factor(df1$PartSpeedCategory)
```

- Inspect the type of variables again

```{r, warning = FALSE, message = FALSE}
str(df1)
```

## Duplicated rows

- Check if there are duplicated rows
  
```{r, warning = FALSE, message = FALSE}
any(duplicated(df1))
```

## Missing values

- Check if there are any missing values
  
```{r, warning = FALSE, message = FALSE}
# in total
sum(is.na(df1))
# per column
colSums(is.na(df1)) 
```

- Investigate the case of missing RT
  
```{r, warning = FALSE, message = FALSE}
df1[!complete.cases(df1[ ,"RT"]), "Word"]
# probably timeouts
# let's check
df1[!complete.cases(df1[ ,"RT"]), "Accuracy"]
# yes, missing RT in combination with error indicates timeout
```

- Investigate the case of missing Frequency
  
```{r, warning = FALSE, message = FALSE}
df1[!complete.cases(df1[ ,"Frequency"]), "Word"]
# all are pseudowords
```

## Check by-participant accuracy

- Create a variable that contains average accuracy per participant

```{r, warning = FALSE, message = FALSE}
df1$PartAcc = ave(df1$Accuracy, df1$Participant)

sort(unique(df1$PartAcc))
```

- Exclude participants with more than 25% errors -- keep only those with accuracy 75% or higher

```{r, warning = FALSE, message = FALSE}
before_p = length(unique(df1$Participant))

df1 = df1[df1$PartAcc >= 0.75, ]

after_p = length(unique(df1$Participant))

NumExcludedPart = before_p - after_p
```

## Check by-item accuracy

- Create a variable that contains average accuracy per item (word)

```{r, warning = FALSE, message = FALSE}
df1$ItemAcc = ave(df1$Accuracy, df1$Word)

sort(unique(df1$ItemAcc))
```

- Exclude items with more than 25% errors -- keep only those with accuracy 75% or higher

```{r, warning = FALSE, message = FALSE}
before_i = length(unique(df1$Word))

df1n = df1[df1$ItemAcc >= 0.75, ]

after_i = length(unique(df1$Word))

NumExcludedItems = before_i - after_i
```

## Let's keep only words and correct responses in the data frame

- only words
```{r, warning = FALSE, message = FALSE}
df1 = df1[df1$Lexicality == "word",]
```

- only correct responses
```{r, warning = FALSE, message = FALSE}
df1 = df1[df1$Error_code == "C",]
```

## Let's transform word frequency, since we know it has a log relationship with RT (Recall the linearity assumption)

 - Later we will also formally check whether the relationship is linear
   but more on that later
   
```{r, warning = FALSE, message = FALSE}
df1$Freq = log(as.numeric(df1$Frequency))

library(ggplot2)
library(patchwork)
g1 = ggplot(df1, aes(x=Frequency, y=RT)) + 
  geom_point() +
  geom_smooth(method = "loess", se = TRUE)

g2 = ggplot(df1, aes(x=Freq, y=RT)) + 
  geom_point() +
  geom_smooth(method = "loess", se = TRUE)

g1 | g2
```

## In addition, continuous predictors should be centered at zero

- We'll see later why (when we discuss the interpretability of the intercept) and it's even better to normalize the values:

```{r, warning = FALSE, message = FALSE}
df1$logFreq = df1$Freq
df1$Freq_z = scale(as.numeric(df1$Freq))
```

- Let's see what we did to Freq_z:

```{r, warning = FALSE, message = FALSE}
ggplot(df1, aes(x=logFreq, y=Freq_z)) + 
  geom_point() +
  geom_smooth(method = "loess", se = TRUE)
```

```{r, warning = FALSE, message = FALSE}
mean(df1$logFreq)
exp(mean(df1$logFreq))
round(mean(df1$Freq_z),5)
```

## Let's also scale the number of senses

```{r, warning = FALSE, message = FALSE}
df1$NoS = as.numeric(df1$NoS)
df1$NoS_z = scale(df1$NoS)
```

## Let's check whether the DV (RT) is normally distributed

```{r, warning = FALSE, message = FALSE}
g1 = ggplot(df1, aes(RT)) + geom_density()
g2 = ggplot(df1, aes(sample=RT)) +
  stat_qq() + stat_qq_line()

g1 | g2
```

- Remove extremely short responses

```{r, warning = FALSE, message = FALSE}
df1 = df1[df1$RT > 300,]
```

## Box-Cox transformation

- Calculates lambda parameter

- Apply the transformation (RT ^ lambda - 1) / lambda

- It boils down to:

  - lambda =1
    - no transformation

  - lambda = 0
    - log transformation
  
  - lambda = -1
    - inverse transformation
    

- Make sure all RT>0
  
```{r, warning = FALSE, message = FALSE}
df1 = df1[df1$RT > 0, ]
```

  - Calculate lambda

```{r, warning = FALSE, message = FALSE}
library(car)
pt = powerTransform(df1$RT)
summary(pt)
lambda = pt$lambda
lambda
```

- Calculate lambda differently

```{r, warning = FALSE, message = FALSE}
powerTransform(df1$RT)
```

  - Apply the transformation

```{r, warning = FALSE, message = FALSE}
df1$RT_bc = bcPower(df1$RT, lambda)
```

  - Inspect the change

```{r, warning = FALSE, message = FALSE}
g3 = ggplot(df1, aes(RT_bc)) + geom_density()
g4 = ggplot(df1, aes(sample=RT_bc)) +
  stat_qq() + stat_qq_line()

(g1 | g2) / (g3 | g4) 
```

- Heuristics, if lambda is closer to -1 than to 0

```{r, warning = FALSE, message = FALSE}
df1$RT_inv = -1000/df1$RT
```

- Plot RT, RT_bc and RT_inv
```{r, warning = FALSE, message = FALSE}

g3 = ggplot(df1, aes(RT_bc)) + geom_density()
g4 = ggplot(df1, aes(sample=RT_bc)) +
  stat_qq() + stat_qq_line()
g5 = ggplot(df1, aes(RT_inv)) + geom_density()
g6 = ggplot(df1, aes(sample=RT_inv)) +
  stat_qq() + stat_qq_line()

(g1 | g2) / (g3 | g4) / (g5 | g6)

```

## Let's create by-item data

- To compute averages by word, we use the aggregate() function
  - The first argument is the variable we are averaging
  - The second argument is a list defining the groups we're averaging over
  - The third argument is the action we're applying

```{r, warning = FALSE, message = FALSE}
nouns = aggregate(df1$RT, list(df1$Word), mean)
```

- Let's name the columns

```{r, warning = FALSE, message = FALSE}
colnames(nouns) = c("Word", "MeanRT")
```

- Now let's take the columns we need from the large data frame

```{r, warning = FALSE, message = FALSE}

nouns_by_item = df1[, c("Word", "Length", "Freq_z", 
                         "FreCategory", "NoS_z", 
                         "PolyLevelCategory")]
```

- Then let's get rid of all the repeated rows, keeping only one row per word

```{r, warning = FALSE, message = FALSE}
nouns_by_item = unique (nouns_by_item)
```

- Finally, let's merge the two new data frames, using the Word column as the key for merging, both in the receiving data frame (nouns_by_item, by.x=) and in the supplying data frame (nouns, by.y=)
  - (This can also be done via the join function — look it up and try to achieve the same thing using one of that function's variants!)
  
```{r, warning = FALSE, message = FALSE}
nouns_by_item = merge(nouns_by_item, nouns, by.x = "Word", by.y = "Word")
```

## Now we have a data frame with average reaction times for words

```{r, warning = FALSE, message = FALSE}
head(nouns_by_item)
```

## Let's create by-participant data

- To compute averages by participant, we use the aggregate() function
  - The first argument is the variable we are averaging
  - The second argument is a list defining the groups we're averaging over
  - The third argument is the action we're applying

```{r, warning = FALSE, message = FALSE}
participants = aggregate(df1$RT, 
                         list(df1$Participant, df1$PolyLevelCategory), mean)
```

- Let's name the columns

```{r, warning = FALSE, message = FALSE}
colnames(participants) = c("Participant", "PolyLevelCategory", "MeanRT")
```

- Now let's take the columns we need from the large data frame

```{r, warning = FALSE, message = FALSE}

nouns_by_participant = df1[, c("Participant", "PolyLevelCategory", "PartSpeed")]
```

- Then let's get rid of all the repeated rows, keeping only one row per participant

```{r, warning = FALSE, message = FALSE}
nouns_by_participant = unique (nouns_by_participant)
```

- Finally, let's merge the two new data frames, using the Participant column as the key for merging
  
```{r, warning = FALSE, message = FALSE}
nouns_by_participant = merge(nouns_by_participant, participants, 
                             by = c("Participant", "PolyLevelCategory"))
```

## Now we have a data frame with average reaction times for words

```{r, warning = FALSE, message = FALSE}
head(nouns_by_participant)
```

## Thank you!



