---
title: 'Day2, Session 3: Linear Mixed Effects Regression - Workflow'
author: 'Dusica Filipovic Durdevic'
date: '2026-09-15'
output:
  revealjs::revealjs_presentation:
    theme: simple
    highlight: pygments
    transition: none
    center: false
    css: custom.css
    slide_level: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## To start, let's load the packages we need

```{r, warning = FALSE, message = FALSE}
library(rms)
library(lme4)
library(languageR)
library(multcomp)
library(emmeans) #library(lsmeans)
library(multcompView)
library(pbkrtest)
library(lmerTest)
library(gbm)
library(randomForest)
library(party) # zameniti sa partykit ili ranger
library(MASS)
library(car)
library(mgcv)
library(itsadug)
library(methods)
library(effects)
library(RePsychLing)
library(knitr)
library(broom)
library(parameters)
library(ggplot2)
library(see)
```



## Then let's check what our working directory is
```{r, warning = FALSE, message = FALSE}
getwd()
```

 - if it isn't the one we want, we set it to be so (it's most convenient for it to be the one containing the data file)
```{r, warning = FALSE, message = FALSE}
setwd("C:/Eowyn/Dokumenti/Obuke/!MOJE_2026")
```

 - let's check whether we succeeded
```{r, warning = FALSE, message = FALSE}
getwd()
```


## Then we load and look at the data

```{r, warning = FALSE, message = FALSE}
dat=read.table("FilipovicDurdevic.Kostic.SerbianPolysemy_short.txt",sep="\t",TRUE)

dim(dat)


colnames(dat)

head(dat)

str(dat)

```

## I like to immediately put whatever will be my categorical predictor into the right format
```{r, warning = FALSE, message = FALSE}
dat$Participant = as.factor(dat$Participant)
dat$Word = as.factor(dat$Word)
```

- let's look again
```{r, warning = FALSE, message = FALSE}
str(dat)
```

## Let's check whether there are Participants who made too many errors
```{r, warning = FALSE, message = FALSE}

sort(tapply(dat$Error_code, dat$Participant, sum)/150)
```

## based on that, let's exclude one Participant
```{r, warning = FALSE, message = FALSE}
dat=dat[dat$Participant!="Participant_57",]

dim(dat)
```


## Let's check whether there are words with a large number of errors
```{r, warning = FALSE, message = FALSE}
sort(tapply(dat$Error_code, dat$Word, sum)/30)
```


## based on that, let's exclude a few words
```{r, warning = FALSE, message = FALSE}
dat=dat[dat$Word!="PISAK",]
dat=dat[dat$Word!="PATENT",]
dat=dat[dat$Word!="SFERA",]
dim(dat)
```
- we excluded about 2% of the data

```{r, warning = FALSE, message = FALSE}
(4500 - 4410)/4500
```



## Let's keep only correct responses
```{r, warning = FALSE, message = FALSE}

dat=dat[dat$Error_code == "1",]
dim(dat)
```
- in total we excluded about 5% of the data

```{r, warning = FALSE, message = FALSE}
(4500 - 4258)/4500
```


## Visual inspection of the RT distribution
```{r, warning = FALSE, message = FALSE}
par(mfrow=c(2,2))
plot(sort(dat$RT))
plot(density(dat$RT))
qqnorm(dat$RT)
par(mfrow=c(1,1))
```


## Let's exclude a few (nonsensically) very fast responses

```{r, warning = FALSE, message = FALSE}
dat=dat[dat$RT>299,]
dim(dat)
```



## Let's check which transformation of RT is recommended
```{r, warning = FALSE, message = FALSE}
powerTransform(dat$RT)
```


## Let's apply the inverse transformation

```{r, warning = FALSE, message = FALSE}
dat$RT=-1000/dat$RT
```


## Let's look at how the new RT is distributed
```{r, warning = FALSE, message = FALSE}
par(mfrow=c(2,2))
plot(sort(dat$RT))
plot(density(dat$RT))
qqnorm(dat$RT)
par(mfrow=c(1,1))
```

## We can also test for normality

```{r, warning = FALSE, message = FALSE}
shapiro.test(dat$RT)
ks.test(jitter(dat$RT),"pnorm",mean(dat$RT),sd(dat$RT))

```


## Let's define right away the inverse function, which we'll need for a nice y-axis when plotting RT
```{r, warning = FALSE, message = FALSE}
trans <- function(x){
  rezultat <- -1000/x
  return(rezultat)
}
```


## Since the main variable in this analysis is H
- we'll use it to illustrate how the trans function is used
- at the same time, here we can already see that this effect might be non-linear

```{r, warning = FALSE, message = FALSE}
ggplot(dat, aes(x=H, y=trans(RT))) + 
  geom_point() +
  geom_smooth(method = "loess", se = TRUE)
```


## Let's log-transform frequency

```{r, warning = FALSE, message = FALSE}
dat$freko=log(dat$freko)
```


## Let's examine collinearity among the predictors

```{r, warning = FALSE, message = FALSE}

collin.fnc(dat[,c("len", "ort", "freko", "fam","conc", "H" )])$cnumber
```

 - The kappa coefficient is huge.
 - 0-6 low
 - 15-29 moderate
 - > 29 high



## Since the Kappa coefficient is catastrophically large, we'll need to resolve the collinearity problem. First, let's look at what the relationship between the variables looks like

```{r, warning = FALSE, message = FALSE}

pairscor.fnc(dat[,c("RT", "len", "ort", "freko", "fam","conc", "H")], hist=TRUE, smooth=TRUE, cex.point=1, col.points="darkgrey")
```

## There are several strategies for resolving collinearity
 - one option is to choose the most influential predictors
 - or to choose one predictor from each group of related predictors
 - or to combine several related predictors into a new score
 - or to run a principal component analysis 


 - to start, let's see how the predictors rank in terms of their influence on the dependent variable

## randomForest

```{r, warning = FALSE, message = FALSE}
rf1 <- randomForest(RT ~ ., data = dat[,c("RT", "len", "ort", "freko", "fam","conc", "H")], importance = TRUE)

varImpPlot(rf1)

importance(rf1)
```



## Gradient Boosting Machines
```{r, warning = FALSE, message = FALSE}
dat.gbm = dat[,c("RT", "len", "ort", "freko", "fam","conc", "H")]

set.seed(125)
gbm1 = summary(gbm1 <- gbm(RT ~
                             len + ort + freko + fam + conc + H,
                           data=dat.gbm, n.trees=500, interaction.depth=3, shrinkage=0.001, cv.folds=5))

gbm1
```



## since in this analysis we want to show that H works over and above the control, already-known variables, i.e. we are not interested in the individual control variables themselves, we choose a different approach — we run a principal component analysis on the set of control variables
```{r, warning = FALSE, message = FALSE}
dat.pca = prcomp(dat[,c("len", "ort","freko","fam", "conc")],
                 center = TRUE,
                 scale. = TRUE)

summary(dat.pca)
```

## let's plot the variance
```{r, warning = FALSE, message = FALSE}
plot(dat.pca, type = "l")
```

## let's plot the percentage of explained variance
```{r, warning = FALSE, message = FALSE}
props = round((dat.pca$sdev ^ 2 / sum(dat.pca$sdev ^ 2)), 3)
barplot(props, col = as.numeric (props > 0.05), 
        xlab = "principal components",
        ylab = "proportion of explained variance")
abline (h = 0.05)
```

 - all of them explain at least 5% of the variance, so we keep them all

## let's look at the loadings
```{r, warning = FALSE, message = FALSE}
dat.pca$rotation
```


## let's write out the coordinates of all words in the space of the new dimensions, creating new, non-collinear control variables
```{r, warning = FALSE, message = FALSE}
dat$pc_1 = dat.pca$x[,1]
dat$pc_2 = dat.pca$x[,2]
dat$pc_3 = dat.pca$x[,3]
dat$pc_4 = dat.pca$x[,4]
dat$pc_5 = dat.pca$x[,5]
```


## Let's examine collinearity in the new set
```{r, warning = FALSE, message = FALSE}
collin.fnc(dat[,c("pc_1","pc_2","pc_3", "pc_4","pc_5", "H" )])$cnumber
```

- this is already more acceptable
 - we see that H is weakly-to-moderately correlated with some of the components
 - however, we'll leave it as is, the Kappa coefficient is moderate, more acceptable
 - and we want to show that H works over and above the control variables
 
```{r, warning = FALSE, message = FALSE}
pairscor.fnc(dat[,c("RT", "pc_1","pc_2","pc_3", "pc_4","pc_5", "H")], hist=TRUE, smooth=TRUE, cex.point=1, col.points="darkgrey")
```





## Let's standardize the predictors
```{r, warning = FALSE, message = FALSE}
dat$trial.z = scale(dat$Trial_order)
dat$freko.z = scale(dat$freko)
dat$len.z = scale(dat$len)
dat$fam.z = scale(dat$fam)
dat$conc.z = scale(dat$conc)
dat$ort.z = scale(dat$ort)
dat$N.z = scale(dat$N)
dat$T.z = scale(dat$T)
dat$H.z = scale(dat$H)
dat$PC1 = scale(dat$pc_1)
dat$PC2 = scale(dat$pc_2)
dat$PC3 = scale(dat$pc_3)
dat$PC4 = scale(dat$pc_4)
dat$PC5 = scale(dat$pc_5)
```
## let's look at collinearity after scaling
```{r, warning = FALSE, message = FALSE}
collin.fnc(dat[,c("PC1","PC2","PC3", "PC4","PC5", "H.z" )])$cnumber
```



## besides the Kappa coefficient, VIF (variance inflation factor) can also be used
 - a VIF greater than 5 or 10 is considered large
```{r, warning = FALSE, message = FALSE}
vif(lm(RT ~ len + ort + freko + fam + conc + H, data = dat))
```


## ##################################################################################################################################

## NOW WE CAN MOVE ON TO FITTING THE MODEL

 - IMPORTANT NOTE
 - STATISTICAL MODELING IS A COMBINATION OF SCIENCE AND SKILL
 - YOU WILL NEVER FULLY MASTER THE TECHNIQUE
 - NEW RECOMMENDATIONS ALWAYS KEEP APPEARING
 - AS DO NEW "RULES" THAT SHOULD BE FOLLOWED
 - AS WELL AS NEW WAYS OF DOING IT



## ##########################################################
## STRATEGY 1
 - we start from the simplest model
 - we add one parameter at a time
 - we compare models
 - if there's no difference, the parameter isn't justified
 - if there is a difference, the parameter is justified, we keep it


## We start with basic models that have only the intercept among the fixed effects
 - just Participants, and just words, as random effects, 
 - we compare these with a model that has both, to be sure of each one's contribution
```{r, warning = FALSE, message = FALSE}
lmer0a <- lmer(RT ~ 1 + 
                 (1 |Participant),
               data=dat, REML=FALSE)
```

```{r, warning = FALSE, message = FALSE}
lmer0b <- lmer(RT ~ 1 + 
                 (1 |Word),	
               data=dat, REML=FALSE)
```

```{r, warning = FALSE, message = FALSE}
lmer0 <- lmer(RT ~ 1 + 
                (1 | Participant) + 
                (1 | Word),	
              data=dat, REML=FALSE)
```

```{r, warning = FALSE, message = FALSE}
anova(lmer0a, lmer0)

```

```{r, warning = FALSE, message = FALSE}

anova(lmer0b, lmer0)
```


## We add trial order to the predictors
 - first we visualize, 
 - then we build the model 
 - and check whether including that parameter is justified

```{r, warning = FALSE, message = FALSE}
ggplot(dat, aes(x=trial.z, y=RT)) +
  geom_point() +
  geom_smooth(method = "lm", se = TRUE) +
  facet_wrap(~Participant) 
```

```{r, warning = FALSE, message = FALSE}

lmer1a <- lmer(RT ~ trial.z + 
                 (1 | Participant) + 
                 (1 | Word),	
               data=dat, REML=FALSE)
```

```{r, warning = FALSE, message = FALSE}
anova(lmer0, lmer1a)
```


## We allow trial order to have a different effect for different Participants

```{r, warning = FALSE, message = FALSE}
lmer1b <- lmer(RT ~ trial.z + 
                 (1 + trial.z | Participant) + 
                 (1 | Word),	
               data=dat, REML=FALSE)
```

```{r, warning = FALSE, message = FALSE}
anova(lmer1b, lmer1a)
```

## We check whether the correlation between the two parameters in the random interactions is justified
 - by creating a model without the correlation
 - and then comparing the fits

```{r, warning = FALSE, message = FALSE}
lmer1bb <- lmer(RT ~ trial.z + 
                  (1 + trial.z || Participant) + 
                  (1 | Word),	
                data=dat, REML=FALSE)		
```

```{r, warning = FALSE, message = FALSE}
anova(lmer1b, lmer1bb)		
```

 -  the correlation is not justified





## Random interaction of trial order with words

```{r, warning = FALSE, message = FALSE}
lmer1c <- lmer(RT ~ trial.z + 
                 (1 + trial.z || Participant) + 
                 (1 + trial.z | Word),	
               data=dat, REML=FALSE)		
```

```{r, warning = FALSE, message = FALSE}
anova(lmer1bb, lmer1c)
```

- not justified, we go back to the previous model, lmer1bb




## Let's add the next predictor
 -  and check whether including it in the model is justified
 -  i.e. whether it contributes to the explained variance
```{r, warning = FALSE, message = FALSE}
lmer2 <- lmer(RT ~ trial.z + PC1 +
                (1 + trial.z || Participant) + 
                (1 | Word),	
              data=dat, REML=FALSE)		
```

```{r, warning = FALSE, message = FALSE}
anova(lmer1bb, lmer2)
```


## We add it to the random effects
```{r, warning = FALSE, message = FALSE}
lmer2a <- lmer(RT ~ trial.z + PC1 +
                 (1 + PC1 | Participant) + (0 + trial.z | Participant) +
                 (1 | Word),	
               data=dat, REML=FALSE)		
```

```{r, warning = FALSE, message = FALSE}
anova(lmer2, lmer2a)
```

## Is the correlation among the random effects justified
```{r, warning = FALSE, message = FALSE}
lmer2b <- lmer(RT ~ trial.z + PC1 +
                 (1 | Participant) + (0 + trial.z | Participant) +
                 (0 + PC1 | Participant) +
                 (1 | Word),	
               data=dat, REML=FALSE)		
```

```{r, warning = FALSE, message = FALSE}
anova(lmer2a, lmer2b)
```

 -  not justified



## We test it for non-linearity
```{r, warning = FALSE, message = FALSE}
lmer2c <- lmer(RT ~ trial.z + poly(PC1,2) +
                 (1 | Participant) + (0 + trial.z | Participant) +
                 (0 + PC1 | Participant) +
                 (1 | Word),	
               data=dat, REML=FALSE)		
```

```{r, warning = FALSE, message = FALSE}
anova(lmer2b, lmer2c)
```

## And so on for every predictor, building up the model
 -  and checking whether the added complication is justified by the data

 -  if we get into a zone where convergence is no longer possible
 -  we simplify the random effects structure
 -  in order to test the fixed effects
 -  and once we've built the final fixed-effects structure
 -  we repeat the testing with the more complex random-effects structure added back in



## Now, for the sake of illustration, let's go ahead and add H

```{r, warning = FALSE, message = FALSE}
lmer3a <- lmer(RT ~ trial.z + PC1 + H.z +
                 (1 | Participant) + (0 + trial.z | Participant) +
                 (0 + PC1 | Participant) +
                 (1 | Word),		
               data=dat, REML=FALSE)		
```

```{r, warning = FALSE, message = FALSE}
anova(lmer2b, lmer3a)
```

## We add it to the random effects
```{r, warning = FALSE, message = FALSE}
lmer3b <- lmer(RT ~ trial.z + PC1 + H.z +
                 (1 + H.z| Participant) + (0 + trial.z | Participant) +
                 (0 + PC1 | Participant) +
                 (1 | Word),		
               data=dat, REML=FALSE)		
```

```{r, warning = FALSE, message = FALSE}
anova(lmer3a, lmer3b)
```

## Is the correlation justified
```{r, warning = FALSE, message = FALSE}
lmer3c <- lmer(RT ~ trial.z + PC1 + H.z +
                 (1 | Participant) + (0 + trial.z | Participant) +
                 (0 + PC1 | Participant) + (0 + H.z | Participant) +
                 (1 | Word),		
               data=dat, REML=FALSE)		
```

```{r, warning = FALSE, message = FALSE}
anova(lmer3b, lmer3c)
```

 -  no, but we're keeping it for now, since this is our key predictor



## We test it for non-linearity
```{r, warning = FALSE, message = FALSE}
lmer3d <- lmer(RT ~ trial.z + PC1 + poly(H, 2) +
                 (1 + H.z| Participant) + (0 + trial.z | Participant) +
                 (0 + PC1 | Participant) +
                 (1 | Word),		
               data=dat, REML=FALSE)		
```

```{r, warning = FALSE, message = FALSE}
anova(lmer3c, lmer3d)
```

 -  it is not non-linear

## THE MODEL HAS ALREADY REPORTED CONVERGENCE PROBLEMS
 - THIS MEANS WE HAVE ALREADY OVERLOADED IT WITH RANDOM EFFECTS...
 - AND WE'VE ONLY JUST BEGUN


## We test the interaction with fixed predictors
```{r, warning = FALSE, message = FALSE}
lmer3e <- lmer(RT ~ trial.z * H.z + PC1 +
                 (1 + H.z| Participant) + (0 + trial.z | Participant) +
                 (0 + PC1 | Participant) +
                 (1 | Word),		
               data=dat, REML=FALSE)		
```

```{r, warning = FALSE, message = FALSE}
anova(lmer3c, lmer3e)
```
 -  this interaction is not justified


```{r, warning = FALSE, message = FALSE}
lmer3f <- lmer(RT ~ trial.z + PC1 * H.z +
                 (1 + H.z| Participant) + (0 + trial.z | Participant) +
                 (0 + PC1 | Participant) +
                 (1 | Word),		
               data=dat, REML=FALSE)		
```

```{r, warning = FALSE, message = FALSE}
anova(lmer3c, lmer3f)
```
 -  this interaction is justified; for now, lmer3f is the best model



## Only once we've shown that a predictor contributes to the explained variance are we allowed to look at whether the coefficient is statistically significant
```{r, warning = FALSE, message = FALSE}
summary(lmer3f)
```



####################################################################################################################		
####################################################################################################################		
## STRATEGY 2 
 -  we start from the most complex, meaningful model
 -  we remove one parameter at a time
 -  we compare models
 -  if there's no difference, that parameter isn't justified
 -  if there is a difference, that parameter is justified, we keep it
```{r, warning = FALSE, message = FALSE}
lmerXa <- lmer(RT ~ trial.z * PC1 * H.z +
                 (1 + trial.z * PC1 * H.z|Participant) + 
                 (1 + trial.z |Word),	
               data=dat, REML=FALSE)
```
 -  first let's try to simplify the random-effects structure
```{r, warning = FALSE, message = FALSE}
summary(rePCA(lmerXa))
```
 -  this will often fail to converge	



## Because of that it's convenient to go back to a simpler random-effects structure to make things easier
- But only for exploratory purposes (to save some time)
  - for the final analyses, we must apply the strategy of starting with the maximum model structure (see previous session)
```{r, warning = FALSE, message = FALSE}
lmerX <- lmer(RT ~ trial.z * PC1 * H.z +
                 (1 |Participant) + 
                 (1 |Word),	
               data=dat, REML=FALSE)
```

## Let's remove one parameter, the triple interaction, and build a new model
```{r, warning = FALSE, message = FALSE}
lmerX1 <- lmer(RT ~ trial.z * PC1 + 
                    trial.z * H.z +
                    PC1 * H.z +
                    (1 |Participant) + 
                    (1 |Word),	
                    data=dat, REML=FALSE)
```

```{r, warning = FALSE, message = FALSE}

anova(lmerX1, lmerX)
```
 -  the triple interaction is not justified


## Now let's remove one two-way interaction at a time
```{r, warning = FALSE, message = FALSE}
lmerX2 <- lmer(RT ~ trial.z * PC1 + 
                 trial.z * H.z +
                 (1 |Participant) + 
                 (1 |Word),	
               data=dat, REML=FALSE)
```

```{r, warning = FALSE, message = FALSE}

anova(lmerX1, lmerX2)
```

 -  the PC1 by H.z interaction is justified

```{r, warning = FALSE, message = FALSE}
lmerX3 <- lmer(RT ~ trial.z * PC1 + 
                 PC1 * H.z +
                 (1 |Participant) + 
                 (1 |Word),	
               data=dat, REML=FALSE)
```

```{r, warning = FALSE, message = FALSE}

anova(lmerX1, lmerX3)
```
 -  the trial.z by H.z interaction is not justified

```{r, warning = FALSE, message = FALSE}
lmerX4 <- lmer(RT ~ 
                 trial.z * H.z +
                 PC1 * H.z +
                 (1 |Participant) + 
                 (1 |Word),	
               data=dat, REML=FALSE)
```

```{r, warning = FALSE, message = FALSE}

anova(lmerX1, lmerX4)
```
 -  the trial.z by PC1 interaction is not justified



## Based on these comparisons, we keep in the model only the parameters whose removal makes a difference

 -  we keep only the interaction that is justified

```{r, warning = FALSE, message = FALSE}
lmerX5 <- lmer(RT ~ trial.z + 
                 PC1 * H.z +
                 (1 |Participant) + 
                 (1 |Word),	
               data=dat, REML=FALSE)
```

```{r, warning = FALSE, message = FALSE}

anova(lmerX1, lmerX5)
```
 -  no difference, so the simpler model lmerX5 performs just as well as the more complex lmerX1




## Let's check whether the remaining non-interaction terms are needed
 -  what about the main effect of trial.z

```{r, warning = FALSE, message = FALSE}
lmerX6 <- lmer(RT ~ 
                 PC1 * H.z +
                 (1 |Participant) + 
                 (1 |Word),	
               data=dat, REML=FALSE)
```

```{r, warning = FALSE, message = FALSE}

anova(lmerX5, lmerX6)
```
 -  it is justified
 - lmerX5 is still the best model


## Now let's bring back the random interactions
-  IF THE MODEL WON'T CONVERGE, FIRST REMOVE THE CORRELATIONS BETWEEN RANDOM EFFECTS
```{r, warning = FALSE, message = FALSE}
lmerY <- lmer(RT ~ trial.z +
                 PC1 * H.z +
                 (1 + trial.z + PC1 * H.z | Participant) + 
                 (1 + trial.z | Word),	
               data=dat, REML=FALSE)
```

```{r, warning = FALSE, message = FALSE}

anova(lmerX6, lmerY)
```
 -  all these random interactions contribute jointly
 -  but
- let's check whether all the random effects are justified
```{r, warning = FALSE, message = FALSE}
summary(rePCA(lmerY))
```


- WE SEE THAT FOR WORDS, THE ENTIRE VARIANCE IS ALREADY EXPLAINED BY A SINGLE PRINCIPAL COMPONENT, WHILE FOR ParticipantS IT IS ALREADY EXPLAINED BY 3 PRINCIPAL COMPONENTS 
- THIS MEANS IT IS JUSTIFIED TO REMOVE 2 PARAMETERS FROM THE BY-Participant RANDOM-EFFECTS STRUCTURE
- IT IS ALSO JUSTIFIED TO REMOVE ONE PARAMETER FROM THE BY-WORD RANDOM-EFFECTS STRUCTURE

- for WORD
- one is enough, we keep the one with the larger variance

- for PARTICIPANT
- 3 are enough
- we remove the two with the smallest variance


## LET'S DISPLAY THE COEFFICIENTS TO DETECT THE COMPONENTS WITH THE SMALLEST VARIANCE, WHICH WE WILL REMOVE

```{r, warning = FALSE, message = FALSE}
lmerY
```

## Let's compare the two models

```{r, warning = FALSE, message = FALSE}
lmerY1 <- lmer(RT ~ trial.z +
                PC1 * H.z +
                (1 + trial.z + H.z | Participant) + 
                (1 | Word),	
              data=dat, REML=FALSE)
```

```{r, warning = FALSE, message = FALSE}
anova(lmerY, lmerY1)
```
- no difference between the two models
- so the added complexity was unnecessary

- the final model is lmerY1



## Only now are we allowed to look at the coefficients
```{r, warning = FALSE, message = FALSE}
summary(lmerY1)
```






## We still need to check whether the effects are non-linear




## Finally, we move on to diagnostics
```{r, warning = FALSE, message = FALSE}
confint(lmerY1, method="Wald")
```
- We see that our effects are always on the same side of zero
- except for the trial-order effect, which isn't very reliable

## Now we will check whether any of the assumptions for applying a linear model have been violated:


- Let's create a column with the predicted DV values:

```{r, warning = FALSE, message = FALSE}
dat$RT.fitted = predict(lmerY1)
```

- Let's create a column with the residuals:

```{r, warning = FALSE, message = FALSE}
dat$RT.res = residuals(lmerY1)
```

- Let's plot the correlation between the fitted values and the residuals
 + To check whether there is homogeneity of variance
 + This should look like a nice "egg"

```{r, warning = FALSE, message = FALSE}
ggplot(dat, aes(x=RT.fitted, y=RT.res)) +
  geom_point() +
  geom_smooth(method = "lm", se = TRUE) 
```

## Let's check whether the residuals are normally distributed

- This should be as close as possible to a straight line:

```{r, warning = FALSE, message = FALSE}
ggplot(dat, aes(sample=RT.res)) +
  stat_qq() + stat_qq_line()
```


## The same thing can also be done like this:

```{r, warning = FALSE, message = FALSE}
par(mfcol=c(1,2))
qqnorm(resid(lmerY1))
plot(fitted(lmerY1), resid(lmerY1))
par(mfcol=c(1,1))
```



## Now we'll remove the points with large residuals

- And check whether they have too much influence on the model
- We refit the model on a subset of points whose residuals are within the range of +/-2.5 sigma

```{r, warning = FALSE, message = FALSE}
lmerY1t <- lmer(RT ~ trial.z +
                 PC1 * H.z +
                 (1 + trial.z + H.z | Participant) + 
                 (1 | Word),	
                 data=dat, 
                 subset=abs(scale(resid(lmerY1)))<2.5,
                 REML=FALSE)
```

- Let's compare the coefficients from the cleaned and the original model
- there shouldn't be drastic changes

```{r, warning = FALSE, message = FALSE}
summary(lmerY1t)
```

## let's repeat the diagnostics 
- things should now look even better

```{r, warning = FALSE, message = FALSE}
par(mfcol=c(1,2))
qqnorm(resid(lmerY1t))
plot(fitted(lmerY1t), resid(lmerY1t))
par(mfcol=c(1,1))
```


## let's visualize the final effect of the main variable

```{r, warning = FALSE, message = FALSE}
ggplot(dat, aes(x=H, y=trans(RT.fitted))) + 
  geom_point() +
  geom_smooth(method = "loess", se = TRUE)


ggsave("H.lmerY1.jpg", width = 5, height = 5)
```


## let's make a table for the paper

```{r, warning = FALSE, message = FALSE}
sjPlot::tab_model(lmerY1t, show.se = TRUE, show.stat = TRUE, dv.labels = c("RT (ms)"))
```


## let's visualize the coefficients

```{r, warning = FALSE, message = FALSE}
plot(model_parameters(lmerY1t))
```
